<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{BiocCheck}
-->

# BiocCheck

`BiocCheck` encapsulates Bioconductor package guidelines
and best practices, analyzing packages and reporting three
categories of issues:

* **REQUIRED** (Equivalent to an `ERROR` in `R CMD check`). This means the package
  is missing something critical and it cannot be accepted into Bioconductor
  until the issue is fixed. (`BiocCheck` will continue past a `required`
  error, thus it is possible to have more than one, but it will end
  with an error code.) 
* **RECOMMENDED** (Equivalent to a `WARNING` in `R CMD check`). We strongly
  encourage that these issues be fixed. In the weeks leading up to a
  Bioconductor release we will ask package authors to fix these issues.
* **NOTE**: Not necessarily something bad, just something we wanted to point
  out. package authors don't need to take action on these.

# Using `BiocCheck`

Most commonly you will use BiocCheck from your operating system command line,
as

    R CMD BiocCheck package

Where `package` is either a directory containing an R package, or a source
tarball (.tar.gz file). R CMD BiocCheck takes options which can be seen by running

    R CMD BiocCheck --help

```{r echo=FALSE}
suppressWarnings(
    output <- system2(file.path(Sys.getenv("R_HOME"), "bin", "R"),
        "CMD BiocCheck --help", stdout=TRUE, stderr=FALSE)
)
cat(paste(output, collapse="\n"))
```

# When should `BiocCheck` be run

Run `BiocCheck` after running `R CMD check`. 

# Installing `BiocCheck`

`BiocCheck` should be installed as follows:

```{r eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite("BiocCheck")
```

This installation process attempts to install a script called `BiocCheck` in the `bin`
directory of your `R` installation. If it fails to do that (most likely due to insufficient
permissions), it will tell you, saying something like:

    Failed to copy the "inst/script/BiocCheck" script to
    /Library/Frameworks/R.framework/Resources/bin
    If you want to be able to run 'R CMD BiocCheck' you'll
    need to copy it yourself to a directory on your PATH,
    making sure it is executable.


You can fix the problem by following these instructions (noting that `R` may
live in a different directory on your system than what is shown above).

Windows users will also see this:

    Windows users need to copy BiocCheck.bat as well.

If you don't have permission to copy these files to the `bin` directory
of your `R` installation, you can, as noted, copy them to any directory
that's in your PATH. For assistance modifying your PATH, see this link
([Windows](http://www.computerhope.com/issues/ch000549.htm)) or 
this one ([Mac/Unix](http://kb.iu.edu/data/acar.html)).

If you manually copy these files to a directory in your PATH that is
not your R bin directory, you'll continue to see the above message 
when (re-)installing `BiocCheck` but you can safely ignore it.

