<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{BiocCheck}
-->

# `BiocCheck`

`BiocCheck` encapsulates Bioconductor package guidelines
and best practices, analyzing packages and reporting three
categories of issues:

* **REQUIRED** (Equivalent to an `ERROR` in `R CMD check`). This means the package
  is missing something critical and it cannot be accepted into Bioconductor
  until the issue is fixed. (`BiocCheck` will continue past a `required`
  error, thus it is possible to have more than one, but it will exit
  with an error code if run from the OS command line.) 
* **RECOMMENDED** (Equivalent to a `WARNING` in `R CMD check`). We strongly
  encourage that these issues be fixed. In the weeks leading up to a
  Bioconductor release we will ask package authors to fix these issues.
* **CONSIDER**: Not necessarily something bad, just something we wanted to point
  out. package authors don't need to take action on these, but they can.
  These are analagous   to `NOTE`s in `R CMD check`, though unlike 
  `NOTE`s, there is no   expectation that they will increase in severity
  over time.

# Using `BiocCheck`

Most commonly you will use BiocCheck from your operating system command line,
as

    R CMD BiocCheck package

Where `package` is either a directory containing an R package, or a source
tarball (.tar.gz file). 

`BiocCheck` can also be run interactively:


```{r eval=FALSE}
library(BiocCheck)
BiocCheck("packageDirOrTarball")
```


R CMD BiocCheck takes options which can be seen by running

    R CMD BiocCheck --help

```{r echo=FALSE}
suppressPackageStartupMessages(library(BiocCheck))
usage()
```

# When should `BiocCheck` be run

Run `BiocCheck` after running `R CMD check`. 

Note that `BiocCheck` is not a replacement for `R CMD check`;
it is complementary. It should be run after `R CMD check` 
completes successfully.

# Installing `BiocCheck`

`BiocCheck` should be installed as follows:

```{r eval=FALSE}
source("http://bioconductor.org/biocLite.R")
biocLite("BiocCheck")
```

This installation process attempts to install a script called `BiocCheck` in the `bin`
directory of your `R` installation. If it fails to do that (most likely due to insufficient
permissions), it will tell you, saying something like:

    Failed to copy the "script/BiocCheck" script to
    /Library/Frameworks/R.framework/Resources/bin. If you want to be able
    to run 'R CMD BiocCheck' you'll need to copy it yourself to a directory
    on your PATH, making sure it is executable. Edit the copied version,
    replacing RSCRIPT_PATH with the full path to Rscript. See BiocCheck
    vignette for more information.

You can fix the problem by following these instructions (noting that `R` may
live in a different directory on your system than what is shown above).

Windows users will also see this:

    Windows users need to copy BiocCheck.bat as well.

If you don't have permission to copy these files to the `bin` directory
of your `R` installation, you can, as noted, copy them to any directory
that's in your PATH. For assistance modifying your PATH, see this link
([Windows](http://www.computerhope.com/issues/ch000549.htm)) or 
this one ([Mac/Unix](http://kb.iu.edu/data/acar.html)).

If you manually copy these files to a directory in your PATH that is
not your R bin directory, you'll continue to see the above message 
when (re-)installing `BiocCheck` but you can safely ignore it.

# Interpreting `BiocCheck` output

## Dependency Checks

* Checks to make sure that there will be no import problems
  if another package imports your package. (`REQUIRED`)
* Reports if it can't figure out how objects were
  initialized (`CONSIDER`).


## Vignette Checks

Can be disabled with `--no-check-vignettes`.

Only run if your package is a software package (as
determined by your [biocViews](#biocViews_Checks)), or if
package type cannot be determined.

* Checks that `vignettes` directory exists. (`REQUIRED`)
* Checks that it contains vignette sources 
  (*.Rmd, *.Rnw, *.Rrst, *.Rhtml, *.Rtex)
  (`REQUIRED`)
* If checking a directory (not a tarball), and vignette sources
  exist in inst/doc, `RECOMMEND` they be removed.
* Report the number of eval=FALSE chunks; if it's more than 50%
  of the total, `RECOMMEND` that more chunks be evaluated.

## Version Checks

* Checks that the version number is valid for a new Bioconductor
  package. This is only done if the `--new-package` option is supplied.
  (`REQUIRED`).
* Checks for a valid version, that format is correct and that 
  version number is appropriate for this version of Bioconductor.
  (`REQUIRED`).

For more information on package versions, see the
[Version Numbering HOWTO](http://www.bioconductor.org/developers/how-to/version-numbering/).

## biocViews Checks

These can be disabled with the `--no-check-bioc-views` option, 
which might be useful when checking non-Bioconductor packages
(since biocViews is a concept unique to Bioconductor).

* Checks that a `biocViews` field is present in the DESCRIPTION file.
  (`RECOMMENDED`).
* Checks for valid views and displays invalid ones. Note that biocViews
  are case-sensitive. (`RECOMMENDED`)
* Checks that all views come from the same parent (one of Software,
  AnnotationData, ExperimentData) (`RECOMMENDED`).

More information about biocViews is available in the
[Using biocViews HOWTO](http://www.bioconductor.org/developers/how-to/biocViews/).

## Build System Compatibility Checks

The Bioconductor Build System (BBS) is our nightly build system and it has
certain requirements. Packages which don't meet these requirements can
be silently skipped by BBS, so it's important to make sure
that every package meets the requirements. All of the following items
are `REQUIRED`.

* Checks to make sure there are no blank lines in the 
  DESCRIPTION file.
* Checks to make sure there is no whitespace in DESCRIPTION
  file field names.
* Checks to make sure that `Package` field of DESCRIPTION file
  matches directory or tarball name.
* Checks to make sure a `Version` field is present in
  the DESCRIPTION file.
* Checks to make sure the DESCRIPTION file has either a `Maintainer`
  field, or a valid `Authors@R` field which resolves to a valid
  `Maintainer`, which consists of:
    * A valid R object of class `person`.
    * Only one person with the `cre` (creator) role.
    * That person must have a syntactically valid email address.

## Unit Test Checks

We strongly believe in unit tests, though we do not at present
require them. For more on what unit tests are, why they are helpful,
and how to implement them, read our
[Unit Testing HOWTO](http://www.bioconductor.org/developers/how-to/unitTesting-guidelines/).

At present we just check to see whether unit tests are present, and if not,
urge you to `CONSIDER` adding them.

## Native Routine Registration Checks

Calls to native (C or Fortran) code entry points should be registered with
R. This is documented in the
[Writing R Extensions](http://cran.r-project.org/doc/manuals/r-release/R-exts.html#Registering-native-routines)
manual. 

If `BiocCheck` detects that your package has native code, but no
entry points have been registered, it will `RECOMMEND` that you 
register them.

## Namespace Import Suggestion Checks

If the package `codetoolsBioC` is installed, `BiocCheck` will run it
to see if it has suggestions for the "Imports" section of your package
NAMESPACE. 

`codetoolsBioC` is an experimental package that is not presently
available via `biocLite()`. It is available from 
[our Subversion repository](https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/codetoolsBioC)
 with the credentials readonly/readonly.

 Output of codetoolsBioC is printed to the screen but `BiocCheck`
 does not label it REQUIRED, RECOMMENDED, or CONSIDER.

## Deprecated Package Checks

 At present, this just looks to see whether your package has a
 dependency on the `multicore` package, and if so, `REQUIRE`s you 
 to remove it. `multicore` does not work on Windows, but the 
 `parallel` package has all of the same functionality and works
 on all operating systems. 

 Note that `parallel` supports two types of parallelism: forking and socket clusters. Forking only works on Windows if the number of cores is set to 1, 
 meaning there is no gain from parallelizing code. Socket clusters
 work on all operating systems.

## Code Parsing Checks

 `BiocCheck` parses the code in your package's R directory, 
 and in evaluated man page and vignette examples to look for various
 symbols, which result in issues of varying severity:

* It's `RECOMMENDED` to replace `T` and `F` with `TRUE` and `FALSE`.
  This is because `T` and `F` are ordinary variables whose value 
  can be altered, leading to unexpected results, whereas the value 
  of `TRUE` and `FALSE` cannot be changed.
* It's `RECOMMENDED` to remove `browser()` calls from code. This 
  function causes the command-line R debugger to be invoked, and 
  should not be used in production code (though it's OK to wrap 
  such calls in a conditional that evaluates to TRUE if some debugging
  option is set).
* `CONSIDER` removing the symbol `<<-` from R code; it's generally
  not a good idea to use it.
* We `RECOMMEND` that you not call `library()` or `require()` on 
  your own package within code in the R directory or in man page
  examples, because it's not necessary. In these contexts, your 
  package is already loaded.

## DESCRIPTION/NAMESPACE consistency checks

`BiocCheck` detects packages that are imported in
NAMESPACE but not DESCRIPTION, or vice versa, and
`RECOMMENDS` fixing them, with an explanation of
how to do so.

## Function length checking

`BiocCheck` prints an informative message about the length
(in lines) of your five longest functions (this includes functions
in your R directory and in evaluated man page and vignette examples).

`BiocCheck` does not assign severity to long functions, but
you may want to consider breaking up long functions into smaller
ones. This is a basic refactoring technique that results
in code that's easier to read, debug, test, and maintain.

## Runnable Examples Checking

`BiocCheck` looks at all man pages which document exported objects
and lists the ones that don't contain runnable examples (either because
there is no `examples` section or because its examples are tagged
with `dontrun`). It's `REQUIRED` that at least 80% of these man
pages have runnable examples. If more than 80% of these pages 
have runnable examples, but some are still missing, `BiocCheck`
lists the missing ones and asks you to `CONSIDER` adding runnable
examples to them.

Runnable examples are a key part of literate programming and
help ensure that your code does what you say it does.

## NEWS checks

`BiocCheck` looks to see if there is a valid NEWS file
(NEWS or NEWS.Rd) either in the 'inst' directory or in 
the top-level directory of your package, and checks whether
it is properly formatted. 

NEWS files are a good way to keep users up-to-date on changes
to your package. Excerpts from properly formatted NEWS
files will be included in Bioconductor release announcements
to tell users what has changed in your package in the last
release.

More information on NEWS files is available in the help topic
`?news`.

## Formatting checks

There is no 100% correct way to format code. There are various
style guides, and these checks adhere to 
[Bioconductor's Style Guide](http://www.bioconductor.org/developers/how-to/coding-style/).

We ask only that you `CONSIDER` adhering to this guide.

In particular, we think it's important to avoid very long lines in code.
Note that some text editors do not wrap text automatically, requiring
horizontal scrolling in order to read it. Also note that R syntax
is very flexible and whitespace can be inserted almost anywhere in
an expression, making it easy to break up long lines.

These checks are run against not just R code, but the DESCRIPTION 
and NAMESPACE files as well as man pages and vignette source files. 
All of these files allow long lines to be broken up.

## Canned Comments check

It can be handy to generate man page skeletons with `prompt()` and/or
RStudio. These skeletons contain comments that look like this:

    %%  ~~ A concise (1-5 lines) description of the dataset. ~~

`BiocCheck` asks you to `CONSIDER` removing such comments.

## bioc-devel Subscription Check

This usually only applies if `BiocCheck` is run on the
Bioconductor build machines, because this step requires
special authorization.

If this authorization is present, `BiocCheck` will 
check to see if the email address in the Maintainer
(or Authors@R) field is subscribed to the bioc-devel
mailing list, and if not, `RECOMMEND`s that you
subscribe. All maintainers should subscribe to this list.
You can subscribe
[here](https://stat.ethz.ch/mailman/listinfo/bioc-devel).
